\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-14 paper.tex

In a prospective ARG, the exhastive recording of all inheritance intervals from generation
to generation leads to a rapid build-up of information, much of which
may not be relevant to the genetic ancestry of the current population. 
Redundant nodes and inheritance intervals can therefore be periodically
``simplified'' to retain only
the ancestry relevant to the current population.
The ``simplify'' algorithm, which prunes away parts of the graph that
are not ancestral to a sample, is the key enabling factor for
this approach to forwards-time simulation and is
described in detail by
\citet{kelleher2018efficient}.
It is a powerful tool for simulation and analysis as it
allows efficient subsetting of ARGs, with the ability to
remove both samples and certain levels of detail about recombination,
as we discuss in the following sections.

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information,
and the conversion process is very similar to
simplifying a prospective gARG with respect to a set of samples.
The process involves two steps, and is illustrated in
Fig.~\ref{fig-ancestry-resolution}.
The first step is straightforward (Fig.~\ref{fig-ancestry-resolution}A,B):
we duplicate the nodes and
edges from the input eARG, and then add inheritance annotations
to the output gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the two outbound edges respectively with the intervals $[0, x)$ and $[x, L)$.
These inheritance interval annotations are clearly in one-to-one
correspondence with the information in the input eARG,
but they are a superset of the
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example
to a gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) containing a breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the sample genomes, \noderef{a}, \noderef{b}, and \noderef{c}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{g}, \noderef{j}, and \noderef{q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have incomplete
bars showing that local ancestry of entirely coalesced regions is omitted.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events are retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure}

Ancestral material is a critical concept in a retrospective ARG,
and refers to the genomic intervals ancestral to the samples
scattered across lineages (the edges of the graph)
at some time in the past. At recombination
events, ancestral material is split between the ancestral lineages.
As genome segments coalesce in a common
ancestor the total amount of ancestral material is reduced, until
eventually, as we traverse rootwards through the graph, all
samples reach a most recent common ancestor at every location
along the genome. The process of transmitting ancestral material
along the edges of an ARG is directly analogous to Hudson's
simulation algorithm for the coalescent with
recombination~\citep{hudson1983testing,kelleher2016efficient}.

Fig.~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
using the ``simplify'' algorithm~\citep{kelleher2018efficient},
and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph by this conversion process.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
retrospective gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.

>>>>

The exhastive recording of full inheritance intervals from generation
to generation leads to a rapid build-up of information, much of which
may not be relevant to the genetic ancestry of the current population.
This redundancy can be periodically removed using the ``simplify'' algorithm,
described in detail by \citet{kelleher2018efficient}. Simplification
allows efficient subsetting of ARGs (removing both samples and certain
levels of detail about recombination, as discussed in Section XXX),
and is also the key enabling factor when incorporating propective ARGs into
forward-time simulation.
The core algorithm involves the concept of
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
the genomic intervals ancestral to a set of samples,
scattered across lineages (the edges of the graph)
at some time in the past. In the next section, we outline how
tracing ancestral material allows the identification and removal of nodes
and inheritance intervals, such that
only genetic ancestry relevant to the current population is retained.

\section*{Converting an eARG to a gARG}
Any eARG can be converted to a gARG without loss of information.
The process is illustrated in Fig.~\ref{fig-ancestry-resolution} (A to B):
we convert all event nodes to genome nodes and add inheritance annotations
to the gARG's edges. If the node is a common ancestor node,
we annotate the single outbound edge with the interval $[0,L)$,
for a genome of length $L$.
If the node is a recombination event with a breakpoint $x$,
we annotate the two outbound edges respectively with the intervals $[0, x)$ and $[x, L)$.
These inheritance interval annotations are clearly in one-to-one
correspondence with the information in the input eARG.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Converting the \citet[][Fig.~1]{wiuf1999recombination} example
to a sample-resolved gARG. (A) The original eARG; square nodes represent events, with
each recombination (red) containing a breakpoint.
(B) The corresponding gARG with breakpoints directly converted to
edges annotated with inheritance intervals.
(C) The sample-resolved gARG resulting from simplifying with respect
to the sample genomes, \noderef{a}, \noderef{b}, and \noderef{c}.
Dashed lines show edges that are
no longer present (in practice, nodes \noderef{g}, \noderef{j}, and \noderef{q} would also be removed).
Coalescence with respect to the sample is indicated by shaded bars, as
in Fig~\ref{fig-arg-in-pedigree}A; nodes \noderef{n}, \noderef{o}, \noderef{p}, \noderef{q} have incomplete
bars showing that local ancestry of entirely coalesced regions is omitted.
Line thickness is proportional to the genomic span of each edge.
Nodes representing recombination events are retained
for clarity, but could be removed by simplification if
desired.
}
\end{figure}

This way of specifying inheritance intervals generates annotations equivalent to
those produced in a prospective gARG, and which is similarly amenable to ``simplification'',
by considering the genetic material that is ancestral to the samples \citep{kelleher2018efficient}.
The outcome of this process, which is
directly analogous to Hudson's simulation algorithm for the coalescent with
recombination~\citep{hudson1983testing,kelleher2016efficient}, is shown in
Fig.~\ref{fig-ancestry-resolution}C. We traverse from the samples rootwards through the graph
keeping track of which material is ancestral to the samples.
At recombination events, ancestral material is split between the ancestral lineages.
As genome segments coalesce in a common
ancestor the total amount of ancestral material is reduced, until
eventually all samples reach a most recent common ancestor at every location
along the genome.

Fig.~\ref{fig-ancestry-resolution}C therefore represents a \emph{sample-resolved} gARG,
% useful to say that we don't *need* to simulate an eARG and convert to a 
% sample-resolved form, but can simulate a sample-resolved gARG directly
of the sort that can be simulated directly by the Hudson algorithm.
This differs in some important ways from its equivalent eARG (Fig.~\ref{fig-ancestry-resolution}A).
Firstly, we can see that some nodes and edges have been removed entirely
from the graph.
The ``grand MRCA'' \noderef{q} is omitted from the
sample-resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between \noderef{g} and \noderef{j} is omitted because the recombination
event at position $5$ (represented by node \noderef{g})
fell in non-ancestral material.
More generally, we can see that the sample resolved
gARG of Fig.~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the
cumulative effects of events over time can be reasoned
about, without first ``replaying'' those events. Many computations
that we wish to perform on an ARG will require resolving
the ancestral material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to perform this resolution of ancestral material once
and to store the result,
whereas the eARG encoding requires us to repeat the process
each time.
